Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-10-24 paper-new/towards-relational-model.tex
small revisions

At this point we might try to weaken from our initial attempt at
providing a poset semantics to a semantics where types are equipped
with merely a reflexive relation, and continue using the
step-insensitive ordering. However, some level of transitivity is
absolutely essential to providing compositional reasoning, and
transitivity is used pervasively in the prior work by New and Licata.
%
The reason that it is essential becomes clear by examining the details of the
graduality proof, so first we expand on how we prove graduality compositionally.
First, we need to extend our compositional semantics of types and terms to a
compositional semantics of \emph{type precision derivations} and \emph{term
precision}.
%
Sticking to a poset-based semantics, if our value types are interpreted as
posets, then we want our precision derivations $c : A \ltdyn A'$ to be
interpreted as a \emph{relation} between $A$ and $A'$. It is also natural to
ask that the relation $c$ interact with the orderings on $A$ and $A'$. More
specifically, if $x' \ltdyn_{A} x$ and $x \mathrel{c} y$, then $x' \mathrel{c}
y$. Similarly, if $x \mathrel{c} y$ and $y \ltdyn_{A'} y'$, then $x \mathrel{c}
y'$. We summarize this requirement by saying $c$ is \emph{downward-closed} in
$A$ and $\emph{upward-closed}$ in $A'$. We will see below why this requirement
is necessary for the graduality proof.

@@

Now that we have an intended model of term precision in the form of
squares, we can see what is required to give a compositional proof of
graduality. Most of the cases of term precision are just congruence
and follow easily. The core of the proof of the graduality property is
in proving the validity of the \emph{cast rules} UpL, UpR, DnL and
DnR.
%
These rules specify a relationship between the semantics of type
precision derivations $c$ and the corresponding casts $\upc c, \dnc c$ (see Section
\ref{sec:GTLC}).
%
To prove graduality compositionally then, we will need to codify in an
extra property on $\sem{c}$ that the relation $\sem{c}$ has some
correspondence with the casts $\sem{\upc c}, \sem{\dnc c}$.
%
What property should we require?

Consider the UpL rule. It states that if

@@

involves an arbitrary other relation $\sem{c'}$! This means we cannot

@@

corresponding to UpL, UpR, DnL, and DnR:

@@

We can use the simpler UpL square to derive the original UpL square by
horizontally composing with the identity square for $c'$:
%
% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMSwxLCJBXzIiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzMsNSwiXFxsdGR5bl97QV8yfSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzUsNCwiYyciLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDMsIlxcdXBje2N9IiwyXSxbMiw0LCJcXGlkIl0sWzEsNSwiXFxpZCIsMl1d
\[\begin{tikzcd}[ampersand replacement=\&]
	{A_1} \& {A_2} \& {A_3} \\
	{A_2} \& {A_2} \& {A_3}
	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{\upc{c}}"', from=1-1, to=2-1]
	\arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
	\arrow["\id"', from=1-2, to=2-2]
	\arrow["\id", from=1-3, to=2-3]
	\arrow["{\ltdyn_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
\end{tikzcd}\]
%
Then to complete the proof, we observe that the composition of $\ltdyn_{A_2}$ with $c'$ is equal to
$c'$, because $c'$ is \emph{downward-closed} under the relation on $A_2$.
An analogous argument shows that we can derive the original UpR square from the simpler version of UpR,
this time using the fact that poset relations are \emph{upward-closed}.
% Thus, we
% obtain the original UpL square by vertically composing the square shown above
% with the square $\id \ltsq{\le_{A_2}\, c'}{c'} \id$ on the bottom.

To recap, we have shown that in order to carry out the graduality proof
compositionally, we need that the relations are closed under the ordering on
each side. In particular, this implies that the ordering relations $\ltdyn_{A}$
must be transitive. Thus, we cannot entirely drop posets from our semantics.

@@

However, we also cannot completely drop step-\emph{insensitivity} from
our model. The graduality property itself is not sensitive to steps,
and in fact there are terms related by term precision that take
differing numbers of steps.
%
The offending step arises precisely from the place where our
definition of $D$ uses a $\later$: the function case.
%
Consider the representability condition corresponding to the DnL rule
for $\iarr \colon \dyntodyn \ltdyn\, \dyn$. This is the square $\sem{\dnc{\iarr}} \ltsq{r({\li D})}{\li\sem{\iarr}} \id$.
%% % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXGxpIEQiXSxbMCwxLCJcXGxpIFUoRCBcXGFyciBcXGxpIEQpIl0sWzEsMCwiXFxsaSBEIl0sWzEsMSwiXFxsaSBEIl0sWzAsMiwiXFxsZXFfe1xcbGkgRH0iLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsIlxcbGlcXGlhcnIiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcZG5je1xcaWFycn0iLDJdLFsyLDMsIlxcaWQiXV0=
%% \[\begin{tikzcd}[ampersand replacement=\&]
%% 	{\li D} \& {\li D} \\
%% 	{\li U(D \arr \li D)} \& {\li D}
%% 	\arrow["{\leq_{\li D}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
%% 	\arrow["{\sem{\dnc{\iarr}}}"', from=1-1, to=2-1]
%% 	\arrow["\id", from=1-2, to=2-2]
%% 	\arrow["\li\sem{\iarr}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
%% \end{tikzcd}\]
For this square to be valid with the lock-step ordering the left and
right hand side would need to take the same number of steps, at least
when the left-hand side does not error.
%
However the right hand side \emph{always} takes $0$ steps, as it is
the identity function, whereas on the left, our definition of
$\sem{\dnc{\iarr}}$ had to take an observable step when its input is a
function: this was inherent to the fact that the function case of the
dynamic type is guarded by a later.

However, observe that we can remedy this particular situation by
replacing the $\id$ on the right hand side by an innocuous function
$\delta^* = (\delta \circ \eta)^\dagger$ that on an input value takes
a single computational step, but is otherwise the identity
function. If we were to ignore computational steps, then this function
\emph{would} be the identity function. We call such a function that is
the identity function except for the introduction of computational
steps a \emph{perturbation}.

>>>>

At this point we might try to weaken our initial attempt at
providing a poset semantics to a semantics where types are equipped
with merely a reflexive relation, and continue using the
step-insensitive ordering. However, some level of transitivity is
absolutely essential to providing compositional reasoning, and
transitivity is used pervasively in the prior work by New and Licata.
%
The reason that it is essential becomes clear by examining the details of the
graduality proof, so first we expand on how we prove graduality compositionally.
First, we need to extend our compositional semantics of types and terms to a
compositional semantics of \emph{type precision derivations} and \emph{term
precision}.
%
Sticking to a poset-based semantics, if our value types are interpreted as
posets, then we want our precision derivations $c : A \ltdyn A'$ to be
interpreted as a \emph{relation} between $A$ and $A'$. It is also natural to
ask that the relation $c$ interact with the orderings on $A$ and $A'$. More
specifically, if $x' \ltdyn_{A} x$ and $x \mathrel{c} y$, then $x' \mathrel{c}
y$. Similarly, if $x \mathrel{c} y$ and $y \ltdyn_{A'} y'$, then $x \mathrel{c}
y'$. We summarize this requirement by saying $c$ is \emph{downward-closed} in
$A$ and $\emph{upward-closed}$ in $A'$. We will see below why this requirement
is necessary for the graduality proof.

@@

Now that we have an intended model of term precision in the form of
squares, we can see what is required to give a compositional proof of
graduality. Most of the cases of term precision are just congruence
and follow easily. The core of the proof of the graduality property is
in proving the validity of the \emph{cast rules} $\upl$, $\upr$, $\dnl$, and
$\dnr$.
%
These rules specify a relationship between the semantics of type
precision derivations $c$ and the corresponding casts $\upc c, \dnc c$ (see Section
\ref{sec:GTLC}).
%
To prove graduality compositionally then, we will need to codify in an
extra property on $\sem{c}$ that the relation $\sem{c}$ has some
correspondence with the casts $\sem{\upc c}, \sem{\dnc c}$.
%
What property should we require?

Consider the $\upl$ rule. It states that if

@@

over an arbitrary other relation $\sem{c'}$! This means we cannot

@@

corresponding to $\upl$, $\upr$, $\dnl$, and $\dnr$:

@@

We can use the simpler $\upl$ square to derive the original $\upl$ square by
horizontally composing with the identity square for $c'$:
%
% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMSwxLCJBXzIiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzMsNSwiXFxsdGR5bl97QV8yfSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzUsNCwiYyciLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDMsIlxcdXBje2N9IiwyXSxbMiw0LCJcXGlkIl0sWzEsNSwiXFxpZCIsMl1d
\[\begin{tikzcd}[ampersand replacement=\&]
	{A_1} \& {A_2} \& {A_3} \\
	{A_2} \& {A_2} \& {A_3}
	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{\upc{c}}"', from=1-1, to=2-1]
	\arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
	\arrow["\id"', from=1-2, to=2-2]
	\arrow["\id", from=1-3, to=2-3]
	\arrow["{\ltdyn_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
\end{tikzcd}\]
%
Then to complete the proof, we observe that the composition of $\ltdyn_{A_2}$ with $c'$ is equal to
$c'$, because $c'$ is \emph{downward-closed} under the relation on $A_2$.
An analogous argument shows that we can derive the original $\upr$ square from the simpler version of $\upr$,
this time using the fact that poset relations are \emph{upward-closed}.
% Thus, we
% obtain the original UpL square by vertically composing the square shown above
% with the square $\id \ltsq{\le_{A_2}\, c'}{c'} \id$ on the bottom.

To recap, we have shown that in order to carry out the graduality proof
compositionally, we need that the relations are closed under the ordering on
each side. In particular, this implies that the ordering relations $\ltdyn_{A}$
must be transitive, so we cannot entirely drop posets from our semantics.

@@

However, we also cannot completely drop step-\emph{insensitivity} from
our model. The graduality property itself is not sensitive to steps,
and in fact there are terms related by term precision that take
differing numbers of steps.
%
The offending step arises precisely from the place where our
definition of $D$ uses a $\later$: the function case.
%
Consider the representability condition corresponding to the $\dnl$ rule
for $\iarr \colon \dyntodyn \ltdyn\, \dyn$. This is the square $\sem{\dnc{\iarr}} \ltsq{r({\li D})}{\li\sem{\iarr}} \id$.
%% % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXGxpIEQiXSxbMCwxLCJcXGxpIFUoRCBcXGFyciBcXGxpIEQpIl0sWzEsMCwiXFxsaSBEIl0sWzEsMSwiXFxsaSBEIl0sWzAsMiwiXFxsZXFfe1xcbGkgRH0iLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsIlxcbGlcXGlhcnIiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcZG5je1xcaWFycn0iLDJdLFsyLDMsIlxcaWQiXV0=
%% \[\begin{tikzcd}[ampersand replacement=\&]
%% 	{\li D} \& {\li D} \\
%% 	{\li U(D \arr \li D)} \& {\li D}
%% 	\arrow["{\leq_{\li D}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
%% 	\arrow["{\sem{\dnc{\iarr}}}"', from=1-1, to=2-1]
%% 	\arrow["\id", from=1-2, to=2-2]
%% 	\arrow["\li\sem{\iarr}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
%% \end{tikzcd}\]
For this square to be valid with the lock-step ordering the left and
right hand side would need to take the same number of steps, at least
when the left-hand side does not error.
%
However, the right hand side \emph{always} takes $0$ steps, as it is
the identity function, whereas on the left, our definition of
$\sem{\dnc{\iarr}}$ had to take an observable step when its input is a
function: this was inherent to the fact that the function case of the
dynamic type is guarded by a later.

Observe that we can remedy this particular situation by
replacing the $\id$ on the right hand side by an innocuous function
$\delta^* = (\delta \circ \eta)^\dagger$ that on an input value takes
a single computational step, but is otherwise the identity
function. If we were to ignore computational steps, then this function
\emph{would} be the identity function. We call such a function that is
the identity function except for the introduction of computational
steps a \emph{perturbation}.
